Inference for Trans-dimensional Bayesian Models with Diffusive Nested Sampling

### 2014-12-07 Documents/Paper/ms.tex
Slight tweaks talking about logl vs logx

Nested Sampling \citep{skilling} overcomes these problems by working with an
alternative sequence of probability distributions, proportional to the prior
but with a hard constraint on the value of the likelihood:
\begin{eqnarray}
p_{l}(\theta) \propto \pi(\theta)\mathds{1}\left[L(\theta) > l\right]\label{eq:constrained_prior}
\end{eqnarray}
The normalising constants of these distributions are given by
\begin{eqnarray}
X(l) &=& \int \pi(\theta)\mathds{1}\left[L(\theta) > l\right] \, d\theta
\end{eqnarray}
which is the amount of prior mass that has likelihood greater than
$l$. As Nested Sampling proceeds, $X$ shrinks geometrically with time, so
the number of iterations required to compress from the slab to the spike is
of order $\log(V_{\rm slab}/V_{\rm spike})$. During this process Nested Sampling
visits states with likelihoods intermediate between those of the slab and the
spike, whereas no choice of $\beta$ in Equation~\ref{eq:annealing} will give
these states substantial probability.

@@

\begin{eqnarray}
p(\theta) &=& \sum_{i=0}^n w_i
\frac{\pi(\theta)\mathds{1}\left[L(\theta) > l_i\right]}{X_i}
\end{eqnarray}
where $\{l_i\}$ is a set of likelihood thresholds or ``levels'',
constructed such that the associated prior mass $X_{i+1}$ of level
$i+1$ is approximately $e^{-1}$ times that of level $i$. The $\{w_i\}$ are a
set of mixture weights, usually chosen to emphasise the higher levels during
the initial stages of a run (while creating levels),
and set to uniform once the desired number of levels $n$ has been created.
See \citet{dnest}
for more details about DNS.

@@

\subsection{Results}
We ran DNS on the simulated data shown in Figure~\ref{fig:sinewave_data}.
The shape of the likelihood function with respect to the prior mass (the
usual likelihood curve associated with Nested Sampling) is shown in
Figure~\ref{fig:sinewaves_likelihood}. Two phase transitions can be seen,
where this curve is concave up. The
first phase transition, at $\log(X) \approx -10$ nats,
separates ``noise-only'' models (where the
entire dataset is accounted for by the noise term) from ``one-sinusoid'' models
where $N=1$. At $\log(X) \approx -35$ nats, a second phase transition separates
$N=1$ models from $N=2$ models. This phase transition causes a double-peaked
feature in the posterior weights (lower panel of
Figure~\ref{fig:sinewaves_likelihood}).

>>>>

Nested Sampling \citep{skilling} overcomes these problems by working with an
alternative sequence of probability distributions, proportional to the prior
but with a hard constraint on the value of the likelihood:
\begin{eqnarray}
p_{l}(\theta) \propto \pi(\theta)\mathds{1}\left[L(\theta) > l\right]\label{eq:constrained_prior}
\end{eqnarray}
The normalising constants of these distributions are given by
\begin{eqnarray}
X(l) &=& \int \pi(\theta)\mathds{1}\left[L(\theta) > l\right] \, d\theta\label{eq:X}
\end{eqnarray}
which is the amount of prior mass that has likelihood greater than
$l$. As Nested Sampling proceeds, $X$ shrinks geometrically with time, so
the number of iterations required to compress from the slab to the spike is
of order $\log(V_{\rm slab}/V_{\rm spike})$. During this process Nested Sampling
visits states with likelihoods intermediate between those of the slab and the
spike, whereas no choice of $\beta$ in Equation~\ref{eq:annealing} will give
these states substantial probability.

@@

\begin{eqnarray}
p(\theta) &=& \sum_{i=0}^n w_i
\frac{\pi(\theta)\mathds{1}\left[L(\theta) > l_i\right]}{X_i}
\end{eqnarray}
where $\{l_i\}$ is a set of likelihood thresholds or ``levels'',
constructed such that the associated prior mass $X_{i+1}$ of level
$i+1$ is approximately $e^{-1}$ times that of level $i$. The $\{w_i\}$ are a
set of mixture weights, usually chosen to emphasise the higher levels during
the initial stages of a run (while creating levels),
and set to uniform once the desired number of levels $n$ has been created.
Since DNS does not sample the posterior distribution, the samples need to be
assigned weights (in a manner similar to importance sampling) to represent
the posterior. See \citet{dnest} for more details about DNS.

@@

\subsection{Results}
We ran DNS on the simulated data shown in Figure~\ref{fig:sinewave_data}.
The (log) likelihood as a function of (log) enclosed prior mass $X$
(Equation~\ref{eq:X}) is plotted in Figure~\ref{fig:sinewaves_likelihood}
This curve is a standard output of Nested Sampling methods and can be used to
gain insight into the structure of the problem. Particularly, when
phase transitions are present, this curve will have concave-up
regions \citep{skilling}.

Two phase transitions can be seen in Figure~\ref{fig:sinewaves_likelihood}. The
first phase transition, at $\log(X) \approx -10$ nats,
separates ``noise-only'' models (where the
entire dataset is accounted for by the noise term) from ``one-sinusoid'' models
where $N=1$. At $\log(X) \approx -35$ nats, a second phase transition separates
$N=1$ models from $N=2$ models. This phase transition causes a double-peaked
feature in the posterior weights (lower panel of
Figure~\ref{fig:sinewaves_likelihood}).
